---
title: "Statistics | HW3"
output: html_notebook
---


# Problem 3: Regression techniques

```{r}
library("glmnet")
ceodata <- read.csv('ceo.csv')
ceodata$X <- NULL
head(ceodata)
```

## Task 1


### 1a
The idea of the lasso regression is to penalize the magnitude of coefficients of features along with minimizing squared residuals.  
In lasso regression coefficients are penalized by adding ${\lambda\|\beta \|_{1}}$ to optimization objective.  
${\lambda}$ is the parameter which balances how much coefficients should be penalized (e.g. if ${\lambda = 0}$ we get simple linear regression).  
Lasso regression performes not only regularization, but also feature selection, because It can set some coefficients to zero.  
Lasso uses L1 norm which is not differentiable at all points (e.g. for 1 dimention It is not differentialbe at origin).  
Therefore, gradient descent can't be used (because gradient is not defined), instead coordinate descent is used as an optimization algorithm.  
  
Main benefit of lasso regression usage comparing to simle linear regression is when we have a huge number of features and want to get sparse solution (features with 0 coefficients can be ignored), because It's can be really hard to implement stepwise selection techniques in high dimensionality cases.

### 1b 
Lasso regression penalize by adding sum of absolute values of coefficients, that depend on the magnitude of each variable.  
It is only reasonably to use this way of penalization when variables are scaled.

### 1c
```{r}
set.seed(100)
# scale and normalize data
normalized.ceodata <- data.frame(scale(ceodata))
normalized.ceodata
x <- model.matrix(salary~.+0, normalized.ceodata)
y <- normalized.ceodata$salary

# cross validation for glmnet, used to choose lambda
cv.lasso <- cv.glmnet(x, y, alpha=1);
plot(cv.lasso)
log(cv.lasso$lambda.min)
coef(cv.lasso, s = "lambda.min")

# plot estimated parameters as functions of lambda
library(plotmo)
lasso <- glmnet(x, y, alpha=1)
plot_glmnet(lasso)
```
Using cross validation, we obtained most efficient lambda = `r cv.lasso$lambda.min`.
As can be seen from plot, using this lambda leads to having 4 non-zero coefficients.

## Task 2


### 2a
```{r}
cleaned.data <- ceodata[ceodata$profits > 0, ]

log.profits <- log(cleaned.data$profits)
log.sales <- log(cleaned.data$sales)

lm.fitted <- lm(log.profits~log.sales)
plot(log.sales, log.profits)
abline(lm.fitted, col='tomato', lwd=1)
```


### 2b
```{r}
y <- log.profits
x <- log.sales

nls.fitted <- nls(y ~ b0 + b1 * x^b2, start=list(b0=0, b1=1, b2=2))
y_hat <- predict(nls.fitted)
summary(nls.fitted)

```

```{r}
# plot linear vs nonlinear regression curves
plot(log.sales, log.profits)
lines(x, predict(lm.fitted), col='tomato', lwd=1)
lines(x, predict(nls.fitted), col="green", lwd=1)
legend("bottomright", legend=c("Nonlinear", "Linear"),col=c("green", "tomato"), lwd=1)
```
We can see on the plot that nonlinear regression curve coincides with linear one almost everywhere except  top right part.  

```{r}
loss.functions = function(x, x.hat)
{
  res = c(mean((x - x.hat) ^ 2), 
          mean(abs(x - x.hat)), 
          mean(abs((x - x.hat) / x )))
  names(res) = c("MSE","MAE", "MAPE")
  return(res);
}
# linear
loss.functions(y, predict(lm.fitted))
# nonlinear
loss.functions(y, predict(nls.fitted))
```
MSE, MAE and MAPE errors for linear and nonlinear regression models.  
We can see that errors for both models are pretty much the same, but nonlinear model is slightly better than linear.  

### 2c
TODO
dependent variable does not have constant variance


## Task 3

### 3a
TODO

### 3b

```{r}
library("readxl")
library("np")
bw1 <- npregbw(log.profits ~ log.sales)
bw1
```
Optimal bandwidth selection method: Least Squares Cross-Validation.

```{r}
bw2 <- npregbw(log.profits ~ log.sales, bwmethod="cv.aic")
bw2
```
Optimal bandwidth selection method: Expected Kullback-Leibler Cross-Validation



```{r}
non.parametric.fitted <- npreg(bws = bw1)
plot(log.sales, log.profits)
lines(log.sales, predict(non.parametric.fitted), col='blue', lwd=1)
```
Here we use optimal bandwidth found by Least Squares Cross-Validation method.

### 3c
```{r}
plot(log.sales, log.profits)
# lines(log.sales, predict(lm.fitted), col='tomato', lwd=1)
lines(log.sales, predict(non.parametric.fitted), col='blue', lwd=1)
lines(log.sales, predict(nls.fitted), col="green", lwd=1)
legend("bottomright", legend=c("Nonlinear", "Nonparametric"),col=c("green", "blue"), lwd=1)
```
We can see on the plot that nonlinear regression curve fits data better that nonlinear one.  


```{r}
# nonlinear
loss.functions(log.profits, predict(nls.fitted))
# nonparametric
loss.functions(log.profits, predict(non.parametric.fitted))

```
MSE, MAE and MAPE errors for nonlinear and nonparametric regression models.  
We can see that nonparametric model are better than nonlinear based on these errors.

## Task 4


### 4a

```{r}
new.ceodata <- ceodata
new.ceodata$salary = NULL
new.ceodata$high.salary <- ifelse(ceodata$salary > 2000, 1, 0)

logit.model <- glm(high.salary ~ ., family=binomial('logit'), data=new.ceodata)
summary(logit.model)
optimal.logit.model <- suppressWarnings(step(logit_model, direction = "backward"))
summary(optimal.logit.model)
```


```{r}
coef(optimal.logit.model)
exp(coef(optimal.logit.model))
```
Assuming fixed totcomp, tenure, profits and assests, odds will change by 1.00004085, so increase a little bit. so probability of getting



```{r}

random_5_rows <- ceodata[sample(nrow(new.ceodata), 5), ]
random_5_rows$high_salary_hat <- predict(optimal.logit.model, newdata = random_5_rows, type = "response")
random_5_rows
```

### 4d
```{r}
threshold <- 0.5
table(optimal.logit.model$y, fitted(optimal.logit.model)>threshold)

library(caret)
conf.matr = confusionMatrix(ifelse(fitted(optimal.logit.model)>threshold,1,0), optimal.logit.model$y, positive = "1")

# sensitivity specificity
conf.matr$byClass[c(1,2)]
```


```{r}
library(pROC)
library("verification")
roc.plot(optimal.logit.model$y, fitted(optimal.logit.model), )
roc = roc(predictor = fitted(optimal.logit.model),response = optimal.logit.model$y)
plot(roc, print.auc=TRUE)
new.threshold = coords(roc, "best",ret="threshold")
new.threshold
```


```{r}
table(optimal.logit.model$y, fitted(optimal.logit.model)>new.threshold)

library(caret)
conf.matr = confusionMatrix(ifelse(fitted(optimal.logit.model)>new.threshold,1,0), optimal.logit.model$y, positive = "1")

# sensitivity specificity
conf.matr$byClass[c(1,2)]

```


## Task 5

### 5a

```{r}
library("tree")
library("rpart")
library("RColorBrewer")
library("rattle")
rpart.ceo = rpart(salary ~ .,data=ceodata,control=rpart.control(cp = 0.01, xval = 10))
fancyRpartPlot(rpart.ceo)
printcp(rpart.ceo)
```


